---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# yenpathy

[![License: Apache 2.0](https://img.shields.io/badge/License-Apache%202.0-blue.svg)](https://opensource.org/licenses/Apache-2.0)
[![Travis build status](https://travis-ci.org/ecohealthalliance/yenpathy.svg?branch=master)](https://travis-ci.org/ecohealthalliance/yenpathy)
[![Codecov test coverage](https://codecov.io/gh/ecohealthalliance/yenpathy/branch/master/graph/badge.svg)](https://codecov.io/gh/ecohealthalliance/yenpathy?branch=master)

## Overview

**yenpathy** is an R package to quickly find *k* shortest paths through a weighted
graph using [Yen's Algorithm](https://dx.doi.org/10.1287/mnsc.17.11.712). This algorithm has numerous
applications in network analysis, such as transportation planning.

There are already comprehensive network analysis packages in R, notably [**igraph**](http://igraph.org/r/) and its tidyverse-compatible interface [**tidygraph**](https://github.com/thomasp85/tidygraph). **yenpathy** complements these
by doing one thing well.

**yenpathy** provides the function `k_shortest_paths()`, which returns *k* shortest paths between two nodes in a graph, ordered from shortest to longest, with length determined by the sum of weights of consecutive edges.

## Installation

Install **yenpathy** from GitHub with [remotes](https://github.com/r-lib/remotes):

```{r eval=FALSE}
library(remotes)
install_github("ecohealthalliance/yenpathy")
```

## Usage

Pass `k_shortest_paths()` a data frame (`graph_df`) with rows representing the edges of a graph, as well as a `from` and `to`, specifying starting and ending nodes in `graph_df`.

```{r, message = FALSE}
library(nycflights13)
library(yenpathy)

small_graph <- data.frame(
  start = c(1, 4, 5, 1, 1, 8, 1, 2, 7, 3),
  end = c(4, 5, 6, 6, 8, 6, 2, 7, 3, 6),
  weight = c(1, 1, 1.5, 5, 1.5, 2.5, 1.5, 0.5, 0.5, 0.5)
)

k_shortest_paths(small_graph, from = 1, to = 6)
```

The function will return a list containing up to `k` (default 1) vectors representing paths through the graph.

```{r}
k_shortest_paths(small_graph,
                 from = 1, to = 6, k = 4)
```

You can also pass an `edge_penalty`, a number which is added to all of the edge weights, effectively penalizing paths consisting of more edges.

```{r}
k_shortest_paths(small_graph,
                 from = 1,
                 to = 6,
                 k = 4,
                 edge_penalty = 1)
```

## Implementation 

The package wraps a C++ implementation of Yen's algorithm created by [Yan Qi](https://github.com/yan-qi). The original C++ code is available on GitHub at [yan-qi/k-shortest-paths-cpp-version](https://github.com/yan-qi/k-shortest-paths-cpp-version).

## Contributing

This project is released with a [Contributor Code of Conduct](https://github.com/ecohealthalliance/yenpathy/blob/master/CONDUCT.md). By participating in this project, you agree to abide by its terms.

Feel free to submit feedback, bug reports, or feature suggestions [here](https://github.com/ecohealthalliance/yenpathy/issues), or to submit fixes as pull requests.
