---
title: "Using Yenpathy"
author: "Toph Allen and Noam Ross, EcoHealth Alliance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{using-yenpathy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: paper.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**yenpathy** is an R package to quickly find *k* shortest paths through a weighted
graph using Yen's Algorithm [@Yen_1971]. This algorithm has numerous
applications in fields from transportation planning to social network analysis.

There are already comprehensive network analysis packages in R, notably
[**igraph**](http://igraph.org/r/) and its tidyverse-compatible interface
[**tidygraph**](https://github.com/thomasp85/tidygraph).**yenpathy** complements
these by doing one thing well.

**yenpathy** provides the function `k_shortest_paths()`, which returns *k* shortest paths between two nodes in a graph, ordered from shortest to longest, with length determined by the sum of weights of consecutive edges.

## Installation

Install yenpathy from GitHub with [remotes](https://github.com/r-lib/remotes):

```{r eval=FALSE}
library(remotes)
install_github("ecohealthalliance/yenpathy")
```

## Usage

To find shortest paths through a network, pass `k_shortest_paths()` a data frame
with rows representing the edges of a graph (`graph_df`), as well as supplying
`from` and `to` nodes in `graph_df`. The function will return a list containing
up to `k` (default 1) vectors representing paths through the graph:

```{r, message = FALSE}
library(yenpathy)

small_graph <- data.frame(
start = c(1, 4, 5, 1, 1, 8, 1, 2, 7, 3),
end = c(4, 5, 6, 6, 8, 6, 2, 7, 3, 6),
weight = c(1, 1, 1.5, 5, 1.5, 2.5, 1.5, 0.5, 0.5, 0.5)
)

k_shortest_paths(small_graph, from = 1, to = 6, k = 4)
```

You can also pass an `edge_penalty`, a number which is added to all of the edge
weights, effectively penalizing paths consisting of more edges.

```{r}
k_shortest_paths(small_graph, from = 1, to = 6, k = 4, edge_penalty = 1)
```

### API

`k_shortest_paths()` takes a data frame as its first argument (`graph_df`). In
place of a data frame, you can provide an [**igraph**](https://igraph.org/r/) or
[**tidygraph**](https://github.com/thomasp85/tidygraph) graph object, which will
be converted to a data frame using `igraph::as_data_frame()`.

By default, the first and second columns are assumed to contain the names of the
the start and end nodes, respectively. These may be either integers or character
vectors. If there is a column named "weights", it is used for edge weights.  You
can change these defaults by passing column numbers or names to the `col_from`,
`col_to`, and `col_weights` arguments.

## An example using flight data

A common application for Yen's algorithm is in transportation planning, when
one wants to find the shortest routes between stops such as airports. Here we use
the `USairports` dataset from the **igraphdata** package to demonstrate how to find
the *k* shortest routes between airports in Bangor, ME (BGR) and Omaha, NE (OMA).

```{r}
library(igraph)
library(igraphdata)
data("USairports")
paths <- k_shortest_paths(USairports, "BGR", "OMA", k = 20, col_weights = "Distance")
paths[1:5]
```

The result shows the airport *nodes*, but we can use the `paths` object to look
up the *edges*, or flights, that make up these itineraries. In this case, we will
select just the carriers with the most seats for each leg, as `USAairports` includes
multiple edges for each carrier operating flights between an airport pair:

```{r}
usa <- as_data_frame(USairports)
itineraries <- 
  lapply(paths, function(.x) {
    n_stops <- length(.x)
    legs <- cbind(.x[1:(n_stops - 1)], .x[2:n_stops])
    flights <- apply(legs, 1, function(.z) {
      fl <- subset(usa, from == .z[1] & to == .z[2])
      fl <- subset(fl, Seats == max(Seats))
    })
    do.call(rbind, flights)
  })
itineraries[1:5]
```

We can calculate the total distance of each itinerary:

```{r}
sapply(itineraries, function(.x) sum(.x[["Distance"]]))
```


### Performance

A number of packages, such as **igraph** and **dodgr**,  provide functions to 
find the shortest path between two nodes on a network, or *all* "simple" (non-looping)
paths between nodes. For the *single shortest* path, these packages can be faster than
**yenpathy**:

```{r}
library(bench)
set.seed(42)
network <- sample_smallworld(1, 20, 4, .05)
bench::mark(
  shortest_paths(network, from = 1, to = 10),
  k_shortest_paths(network, from = 1, to = 10, k = 1),
  check = FALSE
)
```

However, to calculate *k* shortest paths, one generally has to calculate *all*
such paths, and later subset. Doing this with `igraph::all_simple_paths()` can
be considerably slower, even on a very simple network, and impossible due to
memory or time constraints for large, densely-connected graphs.

```{r}
network <- sample_smallworld(1, 8, 3, .05)
bench::mark(
  all_simple_paths(network, from = 1, to = 5, mode = "all"),
  k_shortest_paths(network, from = 1, to = 5, k = 1),
  check = FALSE
)
```

## Implementation 

The package wraps a C++ implementation of Yen's algorithm created by [Yan Qi](https://github.com/yan-qi). The original C++ code is available on GitHub at [yan-qi/k-shortest-paths-cpp-version](https://github.com/yan-qi/k-shortest-paths-cpp-version).

## Contributing

This project is released with a [Contributor Code of Conduct](https://github.com/ecohealthalliance/yenpathy/blob/master/CONDUCT.md). By participating in this project, you agree to abide by its terms.

Feel free to submit feedback, bug reports, or feature suggestions [here](https://github.com/ecohealthalliance/yenpathy/issues), or to submit fixes as pull requests.


## References 
